GVN-Hoist: Hoisting Computations from Branches

### 2016-08-25 hoist.tex
added more missing refs, and added perf improvement for illustration example

GCC recently got code-hoisting which is implemented as part of GVN-PRE: it uses
the set of ANTIC and AVAIL value expressions computed for PRE.  The algorithm
hoists top down in a predecessor where the value is in ANTIC\_IN.  It uses
ANTIC[B] to know what expressions will be computed on every path from B to exit,
and can be computed in B: that is the safety condition.  It uses AVAIL[B] to
subtract out those values already being computed (because they are already
available): this is a redundancy condition.  The cost function is: for each
hoist candidate, if all successors of B are dominated by B, then we know
insertion into B will eliminate all the remaining computations.  It then checks
to see if at least one successor of B has the value available.  This avoids
hoisting it way up the chain to ANTIC.  It also checks to ensure that B has
multiple successors, since hoisting in a straight line is pointless.  The
algorithm continues on down the dominator tree, iterating with PRE until no more
changes.  One advantage of GCC implementation is that it works in sync with the
GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
code-hoisting will hoist them.

@@

\subsection{Illustrative Example}
Code hoisting can also reduce the critical path length of execution in out of
order machines. As more instructios are available at the hoisting point, the
hardware has more instructions to reorder. Following example illustrates how
hoisting can improve performance by exposing more ILP.

\begin{verbatim}
float foo(float d, float min, float max, float a)
{
  float tmin, tmax, inv;

@@

In this program the computations of tmax and tmin are identical to the
computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
depends on inv which depends on a division operation which is generally more
expensive than the addition, subtraction and multiplication operations. The
total latency of computation across each branch is:
$O(div) + 2(O(sub) + O(mul))$
Or, for out of order processors with two add units and two multiply units:
$O(div) + O(sub) + O(mul)$

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float foo(float d, float min, float max, float a)
{
  float tmin, tmax, tmin1, tmax1, inv;

@@

In this code the two subtractions and the division operations can be executed in
parallel because there are no dependencies among them. So the total number of
cycles will be $max(O(div), O(sub)) + O(mul) = O(div) + O(mul)$; since $O(div)$ is
usually much greater than $O(sub)$ \cite{x86,aarch64}

Of course, a partial redundancy elimination pass could just remove the entire
if-block because final operation is an addition (asociative under fast math).

\begin{verbatim}
float foo(float d, float min, float max, float a)
{
  float tmin1, tmax1, inv;

@@

GCC: 23286 has interesting test cases.

@@

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints e.g., intersecting side-effects or, profitability constraints e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximise the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby in cases when
there are no fully redundant instructions. Essentially if,

@@

>>>>

GCC recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of GVN-PRE: it uses the set of ANTIC and AVAIL value expressions computed
for PRE.  The algorithm hoists top down in a predecessor where the value is in
ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
path from B to exit, and can be computed in B: that is the safety condition.  It
uses AVAIL[B] to subtract out those values already being computed (because they
are already available): this is a redundancy condition.  The cost function is:
for each hoist candidate, if all successors of B are dominated by B, then we
know insertion into B will eliminate all the remaining computations.  It then
checks to see if at least one successor of B has the value available.  This
avoids hoisting it way up the chain to ANTIC.  It also checks to ensure that B
has multiple successors, since hoisting in a straight line is pointless.  The
algorithm continues on down the dominator tree, iterating with PRE until no more
changes.  One advantage of GCC implementation is that it works in sync with the
GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
code-hoisting will hoist them.

@@

\subsection{Illustrative Example} \label{subsec:example}
Code hoisting can reduce the critical path length of execution in out of
order machines. As more instructios are available at the hoisting point, the
hardware has more instructions to reorder. Following example illustrates how
hoisting can improve performance by exposing more ILP.

\begin{verbatim}
float fun(float d, float min, float max, float a)
{
  float tmin, tmax, inv;

@@

In this program the computations of tmax and tmin are identical to the
computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
depends on inv which depends on a division operation which is generally more
expensive than the addition, subtraction and multiplication operations. The
total latency of computation across each branch is:
$C(div) + 2(C(sub) + C(mul))$
Or, for out of order processors with two add units and two multiply units:
$C(div) + C(sub) + C(mul)$

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a)
{
  float tmin, tmax, tmin1, tmax1, inv;

@@

In this code the two subtractions and the division operations can be executed in
parallel because there are no dependencies among them. So the total number of
cycles will be $max(C(div), C(sub)) + C(mul) = C(div) + C(mul)$; since $C(div)$ is
usually much greater than $C(sub)$ \cite{x86,aarch64}.

A partial redundancy elimination pass could now remove the entire if-block
because final operation is an addition (asociative under fast math), and the end
result is:

\begin{verbatim}
float fun(float d, float min, float max, float a)
{
  float tmin1, tmax1, inv;

@@

%GCC: 23286 has interesting test cases.

@@

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximize the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints e.g., intersecting side-effects or, profitability constraints e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby in cases when
there are no fully redundant instructions. Essentially if,

@@

The code shown in Section~\ref{subsec:example} is a reduced test that appears in
a hot loop of a proprietary benchmark.  When the expressions are hoisted from
the conditional clauses, the overall performance of that benchmark improves by
$15\%$ on an out-of-order processor due to increased instruction level
parallelism, and better scheduling of the instructions, accommodating for the
long latency of the division operation.
